Supporting Information for "Can we distinguish modes of selective interactions using linkage disequilibrium?

### 2021-03-25 SupportingInformation.Rmd
edits throughout

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "sipreamble.tex"
    extra_dependencies: ["float"]
bibliography: "references.bib"
title: Supporting Information for 
author:
  - Aaron P. Ragsdale^[aaronpeaceragsdale@gmail.com]
date:
  - \today
---

@@

The two-locus diffusion equation with additive selection was first described by
@Kimura1955-qe studied extensively in the 1960s and 70s, including by
@Hill1966-gv and @Ohta1969-ie. The continuous distribution $\psi(x_1, x_2,
x_3)$ of haplotype frequencies in a population, where $x_1$ is the frequency of
$AB$, $x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
Fokker-Planck equation:
\begin{align} \label{eq:diffeq}
\frac{\partial \psi}{\partial \tau} = &
\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
\frac{\partial^2}{\partial x_i \partial x_j} 
\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
  + \frac{\partial}{\partial x_2} D\psi
  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
& - \frac{\gamma_A}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
  \right] \\\nonumber
& -\frac{\gamma_B}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
  \right].
\end{align}
$D$ is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
scaled selection coefficients at the left and right locus, and $\rho$ is the
scaled recombination rate between the two loci. Time $\tau$ is measured in
$2N_e$ generations, and $\nu(\tau)$ is the population size relative
to the ancestral size or some reference size at time $\tau$.

@@

In the method-of-moments approach, instead of solving the differential equation
for $\psi$, we instead integrate both sides of the differential equation
against the multinomial sampling function for a given sampling configuration
$(i, j, k)$. On the left side, we get $\partial_t \Psi_n(i, j, k)$, and on the
right we obtain, after some simple integration by parts and somewhat tedious
simplification, terms for drift, recombination, and selection that can be
written as sparse linear operators of $\Psi_n$. Written compactly, this takes
the form
\begin{equation}
\partial_\tau \Psi_n = 
\frac{1}{2\nu(\tau)}\mathcal{D}_{n}\Psi_n
+ \frac{\rho}{2}\mathcal{R}_{n}\Psi_n
+ \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
+ \mathcal{S}_{n, \mathbf{\gamma}, \mathbf{h}}\Psi_n.
\end{equation}.

Alternatively, we arrive at this same linear system of equations by considering
tracking the expected sampling distribution over $n$ lineages within the full
population and how that changes over time by drawing lineages from one
generation to the next in the style of @Wright1931-wy. Both @Jouganous2017-pq,
for the single-locus SFS, and @Ragsdale2019-nt drew this connection in detail,
so I refer readers to that previous work for a fuller description of those
derivations and discussion. In the next section I repeat the results for
$\mathcal{D}$, $\mathcal{R}$, and $\mathcal{U}$, briefly describe the moment
closure approximation (which is the same as presented in @Ragsdale2019-nt), and
then describe the selection operator $\mathcal{S}$ for selection with
epistasis, both with and without dominance.

@@

If a lineage in our sample of size $n$ recombines in a given generation, which
occurs with probability $nr$, we need to draw an extra lineage from the full
population for it to recombine with. This means we need $\Psi_{n+1}$ in the
previous generation. After drawing that extra lineage, $\Psi_n$ changes as we
draw one of the two recombinant types (each with probability $1/2$$ instead of
the lineage that was chosen to recombine.

@@

We use a jackknife approximation to write the entries of $\Psi_{n+1}$ and
$\Psi_{n+2}$ as linear combinations of entries in $\Psi_n$. The general
strategy is to assume the underlying continuous distribution $\psi(x, y, z)$
can be approximated locally as a quadratic, and then use entries in $\Psi_n$
that are close in frequency to a given entry in $\Psi_{n+l}$ to estimate the
coefficients of that quadratic using the multinomial sampling formula. Then
this quadratic local approximation to $\psi$ can be used to compute
$\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should refer to
section S1.3.5 in the Supporting material for @Ragsdale2019-nt for details.

@@

For a general diploid selection model, the idea is nearly the same, but we need
to draw an extra lineage to determine the fitness of a diploid individual. For
example, the probability that an $AB$ haplotype is paired with an additional
lineage $Ab$ and selected against, and then replaced by an $aB$ haplotype is
\[-ns_{AB/Ab}\frac{i}{n+2}\frac{j+1}{n+1}\frac{k+1}{n}\Psi_{n+2}(i, j+1,
k+1).\] There are now many more possible selective events to consider, but
after accounting for all possible diploid pairs and replacements (90 in total)
and simplifying, we find
\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{n-i-j-k+2}{n+2}\frac{n-i-j-k+1}{n+1}\left(
  s_{AB/ab}i 
  + s_{Ab/ab}j 
  + s_{aB/ab}k
\right)\Psi_{n+2}(i,j,k) \\\nonumber
& + \frac{i+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  - s_{AB/ab}(n + j + k)
)\Psi_{n+2}(i+1, j, k) \\\nonumber
& + \frac{i+2}{n+2}\frac{i+1}{n+1}(
  s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  - s_{AB/AB}(n-i)
)\Psi_{n+2}(i+2, j, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{j+1}{n+2}(
  s_{AB/AB}i
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/Ab}j \\\nonumber & \hspace{20pt}
  + s_{Ab/aB}k
  + s_{Ab/ab}(n-i-j-k)
  - s_{AB/Ab}(2n-i-j)
)\Psi_{n+2}(i+1, j+1, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/aB}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{AB/aB}(2n-i-k)
)\Psi_{n+2}(i+1, j, k+1) \\\nonumber
& + \frac{j+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/ab}i
  + s_{Ab/Ab}j
  + s_{Ab/aB}k \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  -s_{Ab/ab}(n+i+k)
)\Psi_{n+2}(i, j+1, k) \\\nonumber
& + \frac{j+2}{n+2}\frac{j+1}{n+1}(
s_{AB/Ab}i + s_{Ab/aB}k + s_{Ab/ab}(n-i-j-k) - s_{Ab/Ab}(n-j)
)\Psi_{n+2}(i, j+2, k) \\\nonumber
& + \frac{j+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/aB}i
  + s_{Ab/Ab}j
  + s_{Ab/ab}(n-i-j-k) \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{Ab/aB}(2n - j - k)
)\Psi_{n+2}(i, j+1, k+1) \\\nonumber
& + \frac{k+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/aB}i
  + s_{AB/ab}i
  + s_{Ab/aB}j
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  - s_{aB/ab}(n + i + j)
)\Psi_{n+2}(i, j, k+1) \\\nonumber
& + \frac{k+2}{n+2}\frac{k+1}{n+1}(
  s_{AB/aB}i
  + s_{Ab/aB}j
  + s_{aB/ab}(n-i-j-k)
  - s_{aB/aB}(n - k)
)\Psi_{n+2}(i, j, k+2).
\end{align}
Multiplying through by $2N_{ref}$ gives us selection operators in terms of
$\gamma$ instead of $s$.

@@

Loss-of-function (LOF) variants show a dramatic skew toward low-frequency
variants across all human populations (Table \@ref(tab:tajimasD)). Here, using
the folded SFS for synonymous, missense, and LOF mutations across all autosomal
genes, I infer DFEs for missense and LOF mutations independently. I consider a
few different dominance coefficients to explore the effect of the assumed
recessivity of the two classes of mutations.

The standard SFS approach to fitting the DFE involves first inferring a
demographic history for the population using putatively neutral variants (here,
synonymous mutations), and then fixing that demography and fitting a
parameterized function for the distribution of selection coefficients for new
mutations for the selected classes. DFE inference also requires an estimate for
the total mutation rate of the different mutation classes, as much of the
signal for strongly selected mutations comes from observing fewer mutations
than expected given a known mutation rate (with the assumption that selection
purges some fraction of strongly deleterious mutations which are unseen in the
sample). Here, I fit demography and DFEs to the SFS from the Mende in Sierra
Leone (MSL) using *moments* [@Jouganous2017-pq].

I used the mutation model from @Karczewski2020-le to estimate the total
mutation rate across autosomal genes ($uL$, where $u$ is the per-base mutation
rate, and $L$ is the total length of the coding genome). These values were
$(0.1442, 0.3426, 0.0256)$ for synonymous, missense, and LOF mutations,
respectively. Roughly two thirds of new mutations in coding regions are
expected to be missense mutations, while only 5\% of new mutations are LOF. I
fit a demographic model to the synonymous variants, which included a population
expansion in the deeper past and exponential growth in the recent past (Figure
\@ref(fig:msldfe)A). Using the inferred optimal scaled mutation rate,
$\theta=4N_euL$, I estimated $Ne\approx12,300$, and assuming an average
generation time of 29 years I converted the inferred genetic units to physical
units. The best-fit model had a roughly two-fold expansion 400 thousand years
ago, and then exponential growth over the past 20-30 thousand years, with a
current effective size of $\sim 63,000$.

Under this demographic model, I fit a gamma distribution for the distribution
of fitness effects to missense and LOF mutations (Table \@ref(tab:msldfe)). For
each fit, I fixed the scaled mutation rate for each mutation class, so that
$\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}$ and $\theta_{lof} =
\frac{u_{lof}}{u_{syn}} \hat\theta_{syn}$, where values of $u$ were found using
the GNOMAD mutation model [@Karczewski2020-le]. I tested three values for the
dominance coefficient $h$, $0$, $0.2$ and $0.5$. For missense mutations, $h=0$
gave a poor fit to the data, and $h=0.5$ fit best among the three tested
dominance coefficients. For LOF variants, $h=0$ also fit poorly, but $h=0.2$
and $h=0.5$ gave similar likelihoods, highlighting that inferring dominance
using the SFS is poorly constrained. Regardless of the dominance coefficient
assumed, however, the vast majority of LOF variants were inferred to be
strongly deleterious, with only $\sim10\%$ of new mutations having selection
coefficients on the order $1/N_e$ or less.

@@

Multinucleotide mutations can be easily incorporated into the moment system with
a simple adjustment to the mutation operator. Instead of all mutations occurring
independently in haplotypes with mutations already segregating at the other locus,
some fraction of new mutations could instead occur spontaneously and create a new
pair of mutations with initial counts $n_{AB} = 1$ and $n_{ab} = n-1$.

Here, I fit a simple exponential model for the fraction of new mutations at a
given distance that arose through a MNM event, so that $P(MNM | d) =
Ae^{-\lambda d}$, where $d$ is the distance separating pairs of mutations. I
considered all synonymous mutations within genes in the MSL population and used
the same population size history model as inferred in the DFE section above for
a demographic control. This left two parameters to be fit, $A$ and $\lambda$,
which I fit to the binned decay curve of $\sigma_d^1$. I needed to assume an
average per-base recombination rate $r$ across gene regions, and tested a
number of values between $10^{-9}$ and $2\times 10^{-8}$. The optimization was
insensitive to the chosen value or $r$, because the decay of positive LD occurs
rapidly. For any plausible value of $r$, the $\sigma_d^1$ decays to zero well
before distances between pairs have scaled recombination rates $\rho=4 N_e r
d$, and expected statistics for $\rho \ll 1$ vary only negligibly.

In fitting the LD decay of $\sigma_d^1$, the best fit parameters were
$A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
the vast majority of new mutations *pairs* do not occur via MNMs for distances
greater than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs
for very tightly linked loci with distances on the order $0-50$ base pairs. It
is important to note that this does not mean that $10-15\%$ of new mutations
occur via MNMs, since this fraction is contitioned on two mutations occurring
at short distances.

@@

The large confidence intervals for measurements of signed LD could be driven by
either averaging over relatively few observed pairs of mutations, or due to
small sample sizes that make each individual measurement noisy estimates of the
LD for that pair of mutations in the full population. To explore the underlying
cause of measurement uncertainty in the @1000_Genomes_Project_Consortium2015-zq
data, I considered larger sets of samples by combining populations that
consistently cluster together in PCA and UMAP space and have low
differentiation [@Diaz-Papkovich2020-ee]. I took combinations of CEU/GBR,
CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that residual population
structure in these population combinations could alter expected LD statistics
compared to the respective single-population estimates, I was more interested
in the effect that increasing the sample sizes would have on estimated
measurement error.

@@

```{r msldfe, echo=F, tab.align="center"}
library(knitr)
library(kableExtra)
dfes <- read.csv("tables/msl_dfes.txt", sep=";", header=TRUE, check.names=FALSE)
knitr::kable(dfes, "latex", booktabs=T, escape=FALSE,
    linesep = c("", "", "\\addlinespace"),
    caption="DFEs inferred for missense and loss-of-function variants in MSL for varying values of $h$. General patterns are consistent across different chosen values of $h$, although for $h=0$ results in poorer fits for both missense and LOF variants. Columns to the right of the log-likelihood (LL) column show proportions of new mutations with $|s|$ in each given bin.") %>%
kable_styling(latex_options = c("HOLD_position", "repeat_header"))
```

@@

@@

(ref:figmsldfe) **Demography and DFE for MSL.** A demographic model was fit to
the folded synonymous SFS, and DFEs were fit to missense and loss-of-function
SFS. Shown here are DFEs fit with $h=0.5$.

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "sipreamble.tex"
    extra_dependencies: ["float"]
bibliography: "references.bib"
title: Supporting Information for "Can we distinguish modes of selective interactions using linkage disequilibrium?"
author:
  - Aaron P. Ragsdale^[\url{aaronpeaceragsdale@gmail.com}]
date:
  - \today
---

@@

The two-locus diffusion equation with additive selection was first described by
@Kimura1955-qe and studied extensively in the 1960s and 70s [e.g.,
@Hill1966-gv; @Ohta1969-ie]. The continuous distribution $\psi(x_1, x_2, x_3)$
of haplotype frequencies in a population, where $x_1$ is the frequency of $AB$,
$x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
Fokker-Planck equation:
\begin{align} \label{eq:diffeq}
\frac{\partial \psi}{\partial \tau} = &
\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
\frac{\partial^2}{\partial x_i \partial x_j} 
\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
  + \frac{\partial}{\partial x_2} D\psi
  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
& - \frac{\gamma_A}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
  \right] \\\nonumber
& -\frac{\gamma_B}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
  \right].
\end{align}
$D$ is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
(additive) scaled selection coefficients at the left and right locus, and
$\rho$ is the scaled recombination rate between the two loci. Time $\tau$ is
measured in $2N_e$ generations, and $\nu(\tau)$ is the population size relative
to the ancestral size or some reference size at time $\tau$.

@@

In the method-of-moments approach, instead of solving the differential equation
for $\psi$, we instead integrate both sides of the differential equation
against the multinomial sampling function for a given sampling configuration
$(i, j, k)$. On the left side, we get $\partial_t \Psi_n(i, j, k)$, and on the
right we obtain, after some simple integration by parts and somewhat tedious
simplification, terms for drift, recombination, and selection that can be
written as sparse linear operators of $\Psi_n$. Written compactly, this takes
the form
\begin{equation}
\partial_\tau \Psi_n = 
\frac{1}{2\nu(\tau)}\mathcal{D}_{n}\Psi_n
+ \frac{\rho}{2}\mathcal{R}_{n}\Psi_n
+ \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
+ \mathcal{S}_{n, \boldsymbol{\gamma}, \mathbf{h}}\Psi_n.
\end{equation}.

Alternatively, we arrive at this same linear system of equations by tracking
the expected sampling distribution over $n$ lineages within the full population
and how that changes over time by drawing lineages from one generation to the
next in the style of @Wright1931-wy. Both @Jouganous2017-pq, for the
single-locus SFS, and @Ragsdale2019-nt drew this connection in detail, so I
refer readers to those previous papers for a fuller description of those
derivations and discussion. In the next section I simply repeat the results for
$\mathcal{D}$, $\mathcal{R}$, and $\mathcal{U}$, briefly describe the moment
closure approximation (which is the same as presented in @Ragsdale2019-nt), and
then describe the selection operator $\mathcal{S}$ for selection with
epistasis, both with and without dominance.

@@

If a lineage in our sample of size $n$ recombines in a given generation, which
occurs with probability $nr$, we need to draw an extra lineage from the full
population for it to recombine with. This means we need $\Psi_{n+1}$ in the
previous generation. After drawing that extra lineage, $\Psi_n$ changes as we
draw one of the two recombinant types (each with probability $1/2$) instead of
the lineage that was chosen to recombine.

@@

We use a jackknife approximation to write the entries of $\Psi_{n+1}$ and
$\Psi_{n+2}$ as linear combinations of entries in $\Psi_n$. The general
strategy is to assume the underlying continuous distribution $\psi(x_1, x_2,
x_3)$ can be approximated locally as a quadratic, and then use entries in
$\Psi_n$ that are close in frequency to a given entry in $\Psi_{n+l}$ to
estimate the coefficients of that quadratic using the multinomial sampling
formula. Then this quadratically local approximation to $\psi$ can be used to
compute $\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should
refer to section S1.3.5 in the Supporting material for @Ragsdale2019-nt for
details.

@@

For a general diploid selection model, the idea is nearly the same, but we need
to draw an extra lineage to determine the fitness of a diploid individual. For
example, the probability that an $AB$ haplotype is paired with an additional
lineage $Ab$ and selected against, and then replaced by an $aB$ haplotype is
\[-ns_{AB/Ab}\frac{i}{n+2}\frac{j+1}{n+1}\frac{k+1}{n}\Psi_{n+2}(i, j+1,
k+1).\] There are now many more possible selective events to consider, but
after accounting for all possible diploid pairs and replacements (90 in total)
and simplifying, we find

\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{n-i-j-k+2}{n+2}\frac{n-i-j-k+1}{n+1}\left(
  s_{AB/ab}i 
  + s_{Ab/ab}j 
  + s_{aB/ab}k
\right)\Psi_{n+2}(i,j,k) \\\nonumber
& + \frac{i+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  - s_{AB/ab}(n + j + k)
)\Psi_{n+2}(i+1, j, k) \\\nonumber
& + \frac{i+2}{n+2}\frac{i+1}{n+1}(
  s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  - s_{AB/AB}(n-i)
)\Psi_{n+2}(i+2, j, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{j+1}{n+2}(
  s_{AB/AB}i
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/Ab}j \\\nonumber & \hspace{20pt}
  + s_{Ab/aB}k
  + s_{Ab/ab}(n-i-j-k)
  - s_{AB/Ab}(2n-i-j)
)\Psi_{n+2}(i+1, j+1, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/aB}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{AB/aB}(2n-i-k)
)\Psi_{n+2}(i+1, j, k+1) \\\nonumber
& + \frac{j+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/ab}i
  + s_{Ab/Ab}j
  + s_{Ab/aB}k \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  -s_{Ab/ab}(n+i+k)
)\Psi_{n+2}(i, j+1, k) \\\nonumber
& + \frac{j+2}{n+2}\frac{j+1}{n+1}(
s_{AB/Ab}i + s_{Ab/aB}k + s_{Ab/ab}(n-i-j-k) - s_{Ab/Ab}(n-j)
)\Psi_{n+2}(i, j+2, k) \\\nonumber
& + \frac{j+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/aB}i
  + s_{Ab/Ab}j
  + s_{Ab/ab}(n-i-j-k) \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{Ab/aB}(2n - j - k)
)\Psi_{n+2}(i, j+1, k+1) \\\nonumber
& + \frac{k+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/aB}i
  + s_{AB/ab}i
  + s_{Ab/aB}j
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  - s_{aB/ab}(n + i + j)
)\Psi_{n+2}(i, j, k+1) \\\nonumber
& + \frac{k+2}{n+2}\frac{k+1}{n+1}(
  s_{AB/aB}i
  + s_{Ab/aB}j
  + s_{aB/ab}(n-i-j-k)
  - s_{aB/aB}(n - k)
)\Psi_{n+2}(i, j, k+2).
\end{align}
Multiplying through by $2N_{ref}$ gives us selection operators in terms of
$\gamma$ instead of $s$.

@@

Loss-of-function (LOF) variants show a dramatic skew toward low-frequency
variants across all human populations (Table \@ref(tab:tajimasD)). Here, using
the folded SFS for synonymous, missense, and LOF mutations across all autosomal
genes, I inferred DFEs for missense and LOF mutations independently. I
considered a few different dominance coefficients to explore the effect of the
assumed recessivity of the two classes of mutations.

The standard SFS approach to fitting the DFE involves first inferring a
demographic history for the population using putatively neutral variants (here,
synonymous mutations), and then fixing that demography and fitting a
parameterized function for the distribution of selection coefficients for new
mutations for the selected classes. DFE inference also requires an estimate for
the total mutation rate of the different mutation classes, as much of the
signal for strongly selected mutations comes from observing fewer mutations
than expected given a known mutation rate (with the assumption that selection
purges some fraction of strongly deleterious mutations which are unseen in the
sample). Here, I fit demography and DFEs to the folded SFS from the Mende in
Sierra Leone (MSL) using *moments* version 1.1.0 [@Jouganous2017-pq].

I used the mutation model from @Karczewski2020-le to estimate the total
mutation rate across autosomal genes ($uL$, where $u$ is the per-base mutation
rate of a given mutation class, and $L$ is the total length of the coding
genome). These values were $(0.1442, 0.3426, 0.0256)$ for synonymous, missense,
and LOF mutations, respectively. Roughly two thirds of new mutations in coding
regions are expected to be missense mutations, while only 5\% of new mutations
are LOF. I fit a demographic model to the synonymous variants, which included a
population expansion in the deeper past and exponential growth in the recent
past (Figure \@ref(fig:msldemogdfe)A). Using the inferred optimal scaled mutation
rate, $\theta=4N_euL$, I estimated $Ne\approx12,300$, and assuming an average
generation time of 29 years I converted the inferred genetic units to physical
units. The best-fit model had a roughly two-fold expansion 400 thousand years
ago, and then exponential growth over the past 20-30 thousand years, with a
current effective size of $\approx 63,000$.

Under this demographic model, I fit a gamma distribution for the distribution
of fitness effects to missense and LOF mutations (Table \@ref(tab:msldfe)). For
each fit, I fixed the scaled mutation rate for each mutation class, so that
$\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}$ and $\theta_{lof} =
\frac{u_{lof}}{u_{syn}} \hat\theta_{syn}$, where values of $u$ were found using
the GNOMAD mutation model [@Karczewski2020-le]. I tested three values for the
dominance coefficient $h$: $0$, $0.2$ and $0.5$. For missense mutations, $h=0$
gave a poor fit to the data, and $h=0.5$ fit best among the three tested
dominance coefficients. For LOF variants, $h=0$ also fit poorly, but $h=0.2$
and $h=0.5$ gave similar likelihoods, highlighting that inferring dominance
using the SFS is poorly constrained. Regardless of the dominance coefficient
assumed, however, the vast majority of LOF variants were inferred to be
strongly deleterious, with only $\sim10\%$ of new mutations having selection
coefficients on the order $1/N_e$ or less.

@@

MNM events can be easily incorporated into the moment system with a simple
adjustment to the mutation operator. Instead of all mutations occurring
independently in haplotypes with mutations already segregating at the other
locus, some fraction of new mutations could instead occur spontaneously and
create a new pair of mutations with initial counts $n_{AB} = 1$ and $n_{ab} =
n-1$.

Here, I fit a simple exponential model for the fraction of new mutations at a
given distance that arose through a MNM event, so that $P(MNM | d) =
Ae^{-\lambda d}$, where $d$ is the distance separating pairs of mutations. I
considered all synonymous mutations within genes in the MSL population and used
the same population size history model as inferred in the DFE section above for
a demographic control. This left two parameters to be fit, $A$ and $\lambda$,
which I fit to the binned decay curve of $\sigma_d^1$. I needed to assume an
average per-base recombination rate $r$ across gene regions, and tested a
number of values between $10^{-9}$ and $2\times 10^{-8}$. The optimization was
insensitive to the chosen value of $r$, because the decay of positive LD occurs
rapidly. For any plausible value of $r$, $\sigma_d^1$ decays to zero well
before distances between pairs have scaled recombination rates $\rho=4 N_e r d
\sim 1$, and expected statistics for $\rho \ll 1$ vary only negligibly.

In fitting the LD decay of $\sigma_d^1$, the best fit parameters were
$A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
the vast majority of new mutations *pairs* do not occur via MNMs for distances
greater than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs
for very tightly linked loci with distances on the order $0-50$ base pairs. It
is important to note that this does not mean that $10-15\%$ of new mutations
occur via MNMs, since this fraction is conditioned on two mutations occurring
at short distances.

@@

The large confidence intervals for measurements of signed LD could be driven by
either averaging over relatively few observed pairs of mutations, or due to
small sample sizes that make each individual measurement a noisy estimate of
the LD for that pair of mutations in the full population. To explore the
underlying cause of measurement uncertainty in the
@1000_Genomes_Project_Consortium2015-zq data, I considered larger sets of
samples by combining populations that consistently cluster together in PCA and
UMAP space and have low differentiation [@Diaz-Papkovich2020-ee]. I took
combinations of CEU/GBR, CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that
residual population structure in these population combinations could alter
expected LD statistics compared to the respective single-population estimates,
I was more interested in the effect that increasing the sample sizes would have
on estimated measurement error.

@@

```{r msldfe, echo=F, tab.align="center"}
library(knitr)
library(kableExtra)
dfes <- read.csv("tables/msl_dfes.txt", sep=";", header=TRUE, check.names=FALSE)
knitr::kable(dfes, "latex", booktabs=T, escape=FALSE,
    linesep = c("", "", "\\addlinespace"),
    caption="DFEs inferred for missense and loss-of-function variants in MSL for varying values of $h$. General patterns are consistent across different chosen values of $h$, although $h=0$ results in poorer fits for both missense and LOF variants. Columns to the right of the log-likelihood (LL) column show proportions of new mutations with $|s|$ in each given bin.") %>%
kable_styling(latex_options = c("HOLD_position", "repeat_header"))
```

@@

```{r hudson, echo=F, fig.cap="(ref:fighudson)", fig.align="center"}
knitr::include_graphics("figures/hudson_comparison.pdf")
```

(ref:fighudson) The distribution at stationarity of AB haplotype counts in a
sample size of 30, in which we observe 10 $A$ alleles at the left locus, and 10
$B$ alleles at the right locus.

@@

(ref:figmsldfe) **Demography and DFE for MSL.** A demographic model was fit to
the folded synonymous SFS, and DFEs were fit to missense and loss-of-function
SFS. Shown here are DFEs fit with $h=0.5$. See Table \@ref(tab:msldfe) for
inferred best-fit parameters.
